Dynamical properties of the Benguela Current as seen through a cold-spell algorithm.

### 2019-04-15 Write_up/Writing/dissertation.Rmd
updated_commit

For this analysis, the seawater temperature data, obtained from the seven different sources, were combined and formatted into standardized comma delineated value (CSV) files which allowed for a fixed methodology to be used across the entire dataset. Prior to data analysis, all data points exceeding 35°C and/or below 0°C were removed as these were considered as outliers. These data points were then changed to NA (not available) so as to not interfere with analysis. All analyses were conducted in R software version 3.4.2 (http://www.r-project.org/). The data used within this study and comprehensive script used for data analyses, and production of figures can be found at https://github.com/AmierohAbrahams/MASTERSPROJECT.

@@

### Chapter 2 Upwelling

##Introduction

Upwelling events are known to have a large influence on the abundance, diversity and distribution of marine organisms and critically affect marine ecosystem processes. The changes in the intensity and duration of upwelling events have proven to cause large changes in ecosystem functioning [@wang]. Globally, there are four major eastern boundary upwelling systems (EBUS). Each of these systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun ; @pauly]. Eastern boundary upwelling regions only cover 1% of the worlds ocean surface, but are among the most productive regions of the world. This is because these regions are influenced by upwelling events. Upwelling is defined as a process whereby cold, nutrient rich waters are pushed to the surface as a result of alongshore equatorial winds. The South African coastline, being approximately 3000km in length, is unique in that it is bordered by the Agulhas current along the east coast and the Benguela current, an eastern boundary upwelling system, along the west coast. These regions display noticeable differences in seawater temperatures in comparison to each other, primarily due to the influences of the neighbouring ocean currents (Bolton et al. 2004; Mead et al. 2013; Schlegel and Smit 2016). The Benguela Current is driven by the anticyclone high pressures systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region (Shannon 2006; Hutchings et al. 2009). The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast (Lüning 1990; Lutjeharms, Cooper, and Roberts 2000; Hutchings et al. 2009; Schlegel et al. 2017).

Over the past three decades, climate change as a result of anthropogenic activities has established global concern and is shown to have negatively affected marine and terrestrial ecosystems. The consequences of these changes are also shown to have direct effects on the behaviour, survival and functioning of marine organisms, thus ultimately affecting species trophic levels and integration [@doney]. According to the Bakun hypothesis, an increase in greenhouse gases, will result in an increase in day time warming and night time cooling, this ultimately causes an increase in temperature gradients and this further results in a strong pressure gradient. Pressure gradients are known to affect wind patterns and wind patterns subsequently control the intensity and duration of upwelling events  [@hsieh; @mote; @lima].

@@

Global warming is increasing at an extremely high rate. This will result in an increase in greenhouse gases which in turn will impact the ocean and terrestrial life. The process of upwelling is one such factor that is predicted to change with an increase in global warming. As prreviously mentioned, changes in temperature results in changes in pressure gradients, these pressure gradients control wind patterns which ultimately controls the intensity and duration of upwelling. To assess the changes within the upwelling events, various datasets would be used. These are satellite sea surface temperatures (SST), in situ collected seawater temperatures, wind data as well as satellite obtained chlorophyll-a data.

@@

Wind action was an important variable in this study as it was hypothesised to exhibit a direct
influence on seawater temperatures (Sinnett and Feddersen 2014) and a direct influence on the intensity and duration of upwelling events; consequently, wind action was investigated for its impact on upwelling at specific sites along the west coast of South Africa. Wind data was obtained from the South African Weather Service (SAWS), and were provided at three-hour resolutions. Specific wind characteristics were measured, namely, wind direction (dirw) and wind speed (spw).  A resolution of 200m was modelled at both 7 and 15m isobaths.

@@

@@

In order to compare the impact of wind action on the intensity and duration of upwelling, the appropriate sites along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, Firstly the full dataset was narrowed to only those sites that could be adequately compared. To do this, the filter () function in R was used. This function selects all the sites with a time series of greater than 10 years. Once the sites were minimised, one years worth of data was selected from the SACTN data between the years 2014 – 2015.  This yield a total of  seven sites along the west coast of South Africa. Hereafter, each of the remotely sensed SST datasets obtained were matched with the SACTN dataset. This was done for two reasons, to find the exact same points at which temperature was collected for the SACTN sites and secondly, to confirm that the dates and time that temperature were collected correspond accordingly.

>>>>

For this analysis, the seawater temperature data, obtained from the seven different sources, were combined and formatted into standardized comma delineated value (CSV) files which allowed for a fixed methodology to be used across the entire dataset. Prior to data analysis, all data points exceeding 35°C and/or below 0°C were removed as these were considered as outliers. These data points were then changed to NA (not available) so as to not interfere with analysis. All analyses were conducted in R software version 3.4.2 (http://www.r-project.org/). The data used within this study and comprehensive script used for data analyses, and production of figures can be found at https://github.com/AmierohAbrahams/MastersProject.

@@

### Chapter 2 Upwelling - Paper 1

## Introduction - Paper 1

Upwelling events are known to have a large influence on the abundance, diversity and distribution of marine organisms and critically affect marine ecosystem processes. The changes in the intensity and duration of upwelling events have proven to cause large changes in ecosystem functioning [@wang]. Globally, there are four major eastern boundary upwelling systems (EBUS). Each of these systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun ; @pauly]. Eastern boundary upwelling regions only cover 1% of the worlds ocean surface, but are among the most productive regions of the world. This is because these regions are influenced by upwelling events. Upwelling is defined as a process whereby cold, nutrient rich waters are pushed to the surface as a result of alongshore equatorial winds. The South African coastline, being approximately 3000km in length, is unique in that it is bordered by the Agulhas current along the east coast an d the Benguela current, an eastern boundary upwelling system, along the west coast. These regions display noticeable differences in seawater temperatures in comparison to each other, primarily due to the influences of the neighbouring ocean currents (Bolton et al. 2004; Mead et al. 2013; Schlegel and Smit 2016). The Benguela Current is driven by the anticyclone high pressures systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region (Shannon 2006; Hutchings et al. 2009). The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast (Lüning 1990; Lutjeharms, Cooper, and Roberts 2000; Hutchings et al. 2009; Schlegel et al. 2017).

Over the past three decades, climate change as a result of anthropogenic activities has established global concern and is shown to have negatively affected marine and terrestrial ecosystems. The consequences of these changes are also shown to have direct effects on the behaviour and functioning of marine organisms, thus ultimately affecting species trophic levels and integration [@doney]. According to the Bakun hypothesis, an increase in greenhouse gases, will result in an increase in day time warming and night time cooling, this ultimately causes an increase in temperature gradients and this further results in a strong pressure gradient. Pressure gradients are known to affect wind patterns and wind patterns subsequently control the intensity and duration of upwelling events  [@hsieh; @mote; @lima].

@@

Global warming is increasing at an extremely high rate. This will result in an increase in greenhouse gases which in turn will impact the ocean and terrestrial life. The process of upwelling is one such factor that is predicted to change with an increase in global warming. As prreviously mentioned, changes in temperature results in changes in pressure gradients, these pressure gradients control wind patterns which ultimately controls the intensity and duration of upwelling. To assess the changes within the upwelling events, various datasets would be used. These are satellite sea surface temperatures (broad scale), in situ collected seawater temperatures (local scale), wind data as well as satellite obtained chlorophyll-a data. Having assessed the eastern boundry upwelling system within a context that comprised of broad and local-scale features permitted us to assess how upwelling varied between sites located closer to the upwelling centres and furthur away from the upwelling centre. Specifically, we aim to assess whether or not it is possible to identify the upwelling signal throughout the various datasets and here we examine how far offshore the upwelling signal is present in the higher resolution data and how far offshore the SACTN signal extend. We hypothesized that (i) wind action may influence the intensity and duration of upwelling; (ii) Remotely sensed SST data will represent a similar event frequency and duration; (iii) Sites located closer to upwelling centres displays distinct variation in the intensity and duration of upwelling events compared to sites furthur away from the upwelling centeres.

@@

@@

An obvious previledge exist when making using of in situ data over satellite data as they provide a more accurate representation of the thermal properties closer to the coast where satellite temperatures do not capture temperature properties well. Smit et al (2013) have shown that satellite data collected along the South African coastline represent a warm biases reaching as high as 6°C over in situ temperatures within the nearshore environment.

Inorder to create a time series of each of the remotely sensed SST data representing and also comparing the broad-scale temperatures against the in situ time series, shore-normal transects were extended from the in situ sites a long the west coast of South Africa. Temperature values from the various satellite datasets were extracted at multiple distances from the coast. These distances are 15 x 15 km, 25 X 25 km and 50 X 50 km pixels along the transect (Figure 1). The daily sea surface temperatures obtained from the various remotely sensed datasets within each of the transects were initiated inorder to produce offshore, broad-scale time series that could accurately be used as a dataset to be compared against the in situ time series.

Wind action was an important variable in this study as it was hypothesised that wind direction and wind speed may have a direct influence on the intensity and duration of upwelling events; consequently, wind action was investigated for its impact on upwelling at specific sites along the west coast of South Africa. Wind data was obtained from the South African Weather Service (SAWS), and were provided at three-hour resolutions. Specific wind characteristics were measured, namely, wind direction (dirw) and wind speed (spw).

@@

At a broad scale, the South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions, each with contrasting temperatures [@smit, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings]. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflect a distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. In order to examine upwelling patterns between sites, the appropriate sites along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites that could be adequately compared. Once the sites were minimised, one years worth of data was selected from the SACTN data between the years 1992 – 2015.  This yield a total of six sites along the west coast of South Africa. Hereafter, each of the remotely sensed SST datasets were matched to the SACTN dataset. This was done for two reasons, to find the same points at which temperature was collected for the SACTN sites and secondly, to confirm that the dates and time that temperature were collected correspond accordingly.

## Detecting the intensity and duration of upwelling
